---
title: "exercise-11"
output: html_document
date: "2024-03-25"
---

# Exercise 11: Practice ANOVA

```{r}
library(tidyverse)
library(dplyr)
library(mosaic)
library(readr)
library(skimr)
library(broom)
library(car)
```

## Preliminaries

```{r}
#Load "AVONETdataset1.csv" dataset as a tibble
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
(d <- read_csv(f,col_names = TRUE))
```

```{r}
#Winnow the dataset to include only the following variables: Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, and Range.Size

(d <- d |> select("Species1","Family1","Order1","Beak.Length_Culmen","Beak.Width","Beak.Depth","Tarsus.Length","Wing.Length","Tail.Length","Mass","Habitat","Migration","Trophic.Level","Trophic.Niche","Min.Latitude","Max.Latitude","Centroid.Latitude","Primary.Lifestyle","Range.Size"))
```

```{r}
#Exploratory data analysis
(skim <- skim(d))
```

In the winnowed dataset, d...

...the following variables (total = 12) are numeric:

```{r}
skim_num <- skim |> filter(skim_type == "numeric")
(skim_num$skim_variable)
```

...and the following variables (total = 7) are 
categorical:

```{r}
skim_char <- skim |> filter(skim_type == "character")
(skim_char$skim_variable)
```

## Challenge 1: One-Factor ANOVA and Inference

### Step 1

Make boxplots of log(Mass) in relation to Trophic.Level and Migration behavior 
type. For each plot, drop from the visualization all species records where the 
categorical variable of interest is missing from the dataset.

```{r}
#Converting the variable Migration from class numeric to either being classified as a factor or as a character (string) variable.

d$Migration <- as.character(d$Migration)

#Checking that this worked
(class(d$Migration))
```

```{r}
#Dropping all species records where Trophic.Level is missing from the dataset
d2 <- d |> drop_na("Trophic.Level")

#Box plot of log(Mass) in relation to Trophic.Level 
p1 <- ggplot(data = d2, aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot(na.rm = TRUE) + 
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("log(Mass)") + 
  xlab("Trophic Level") 
(p1)
```

```{r}
#Dropping all species records where Migration is missing from the dataset
d3 <- d |> drop_na("Migration")

#Box plot of log(Mass) in relation to Migration behavior type
p1 <- ggplot(data = d3, aes(x = Migration, y = log(Mass))) +
  geom_boxplot(na.rm = TRUE) + 
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("log(Mass)") + 
  xlab("Migration Behavior") 
(p1)
```

### Step 2

Run linear models using the lm() function to look at the relationship between log(Mass) and Trophic.Level and between log(Mass) and Migration.

```{r}
#Linear model looking at relationship between log(Mass) and Trophic.Level
(lm1 <- lm((log(Mass)) ~ Trophic.Level, data = d2))
```

```{r}
(summary(lm1))
```

```{r}
#Linear model looking at relationship between log(Mass) and Migration
(lm2 <- lm((log(Mass)) ~ Migration, data = d3))
```

```{r}
(summary(lm2))
```

