---
title: "exercise-11"
output: html_document
date: "2024-03-25"
---

# Practice ANOVA

## Preliminaries

```{r}
library(tidyverse)
library(dplyr)
library(mosaic)
library(readr)
library(skimr)
library(broom)
library(car)
library(manipulate)
library(ggplot2)
library(infer)
```

```{r}
#Load "AVONETdataset1.csv" dataset as a tibble
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
(d <- read_csv(f,col_names = TRUE))
```

```{r}
#Winnow the dataset to include only the following variables: Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, and Range.Size

(d <- d |> select("Species1","Family1","Order1","Beak.Length_Culmen","Beak.Width","Beak.Depth","Tarsus.Length","Wing.Length","Tail.Length","Mass","Habitat","Migration","Trophic.Level","Trophic.Niche","Min.Latitude","Max.Latitude","Centroid.Latitude","Primary.Lifestyle","Range.Size"))
```

```{r}
#Exploratory data analysis
(skim <- skim(d))
```

In the winnowed dataset, d...

...the following variables (total = 12) are numeric:

```{r}
skim_num <- skim |> filter(skim_type == "numeric")
(skim_num$skim_variable)
```

...and the following variables (total = 7) are categorical:

```{r}
skim_char <- skim |> filter(skim_type == "character")
(skim_char$skim_variable)
```

## Challenge 1: One-Factor ANOVA and Inference

### Step 1

#### Make boxplots of log(Mass) in relation to Trophic.Level and Migration behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset.

```{r}
#Converting the variable Migration from class numeric to either being classified as a factor or as a character (string) variable.

d$Migration <- as.factor(d$Migration)

#Checking that this worked
(class(d$Migration))
```

```{r}
#Dropping all species records where Trophic.Level is missing from the dataset
d2 <- d |> drop_na("Trophic.Level")

#Box plot of log(Mass) in relation to Trophic.Level 
p1 <- ggplot(data = d2, aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot(na.rm = TRUE) + 
  geom_jitter(alpha = 0.05)
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("log(Mass)") + 
  xlab("Trophic Level") 
(p1)
```

```{r}
#Dropping all species records where Migration is missing from the dataset
d3 <- d |> drop_na("Migration")

#Box plot of log(Mass) in relation to Migration behavior type
p2 <- ggplot(data = d3, aes(x = Migration, y = log(Mass))) +
  geom_boxplot(na.rm = TRUE) + 
  geom_jitter(alpha = 0.05) +
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("log(Mass)") + 
  xlab("Migration Behavior") 
(p2)
```

### Step 2

#### Run linear models using the lm() function to look at the relationship between log(Mass) and Trophic.Level and between log(Mass) and Migration.

```{r}
#Linear model looking at relationship between log(Mass) and Trophic.Level
(lm1 <- lm((log(Mass)) ~ Trophic.Level, data = d2))
```

```{r}
(summary(lm1))
```

```{r}
#Linear model looking at relationship between log(Mass) and Migration
(lm2 <- lm((log(Mass)) ~ Migration, data = d3))
```

```{r}
(summary(lm2))
```

Both linear models are significant. Log(Mass) is significantly associated with Trophic.Level, and log(Mass) is significantly associated with Migration category.

For the Migration model, **Migration category 1** is the **reference level**. **Migration category 2** and **Migration category 3** both differ significantly from Migration category 1.

```{r}
#Releveling and assessing differences among the remaining pair of Migration categories.

d4 <- d |>
  mutate(Migration = relevel(Migration, ref = "3"))

lm3<- lm(log(Mass) ~ Migration, data = d4)
(summary(lm3))
```

After releveling, **Migration category 3** is the **reference level**. Migration category 2 also differs significantly from Migration category 3.


### Step 3

#### Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another.

```{r}
#One-Factor ANOVA
maov1 <- aov((log(Mass)) ~ Migration, data = d3)

#post-hoc Tukey Honest Significant Differences test
(posthoc <- TukeyHSD(maov1, which = "Migration", conf.level = 0.95))
```

Looking at the adjusted p-values, all three Migration categories significantly differ from each other.

### Step 4

#### Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(Mass) in relation to Trophic.Level and calculate a p value for your original F statistic.

```{r}
#Results for F statistic and p-value for Omnibus F test
(original.F <- aov(log(Mass) ~ Trophic.Level, data = d2) |>
  tidy() |>
  filter(term == "Trophic.Level"))
```

```{r}
#Creating new variable, logMass
d2 <- d2 |> mutate(logMass = log(Mass))

#Permutation approach to inference using {infer}
permuted.F <- d2 |>
  specify(logMass ~ Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

#Histogram
visualize(permuted.F) + 
  shade_p_value(obs_stat = original.F$statistic, direction = "greater")
```

```{r}
#Calculating p-value for original F statistic
p.value <- permuted.F |> get_p_value(obs_stat = original.F$statistic, direction = "greater")
(original.F$p.value)
```


## Challenge 2: Data Wrangling, One and Two-Factor ANOVA

### Step 1

#### Create the following two new variables and add them to AVONET dataset: Relative beak length and Relative tarsus length.

```{r}
#Relative beak length, calculated as the residual of log(Beak.Length_Culmen) on log(Mass)
d$relBeak.Length <- resid(lm(formula = log(Beak.Length_Culmen) ~ log(Mass), data=d))

#Relative tarsus length, calculated as the residual of log(Tarsus.Length) on log(Mass)
d$relTarsus.Length <- resid(lm(formula = log(Tarsus.Length) ~ log(Mass), data=d))
```


### Step 2

#### Make a boxplot or violin plot of your new relative tarsus length variable in relation to Primary.Lifestyle and of your new relative beak length variable in relation to Trophic.Niche

```{r}
#Boxplot for relative tarsus length in relation to Primary.Lifestyle
(p3 <- ggplot(data = d |> filter(!is.na(Primary.Lifestyle)), aes(x = Primary.Lifestyle, y = relTarsus.Length)) +
  geom_boxplot() +
   theme(axis.text.x = element_text(angle = 90)))
```

```{r}
#Boxplot for relative tarsus length in relation to Trophic.Niche
(p4 <- ggplot(data = d |> filter(!is.na(Trophic.Niche)), aes(x = Trophic.Niche, y = relTarsus.Length)) +
  geom_boxplot() +
   theme(axis.text.x = element_text(angle = 90)))
```


### Step 3

#### Run an ANOVA analyses to look at the association between geographic range size and the variable Migration.

```{r}
#Looking at distribution of Range.Size
ggplot(data = d, aes(x = Range.Size)) +
    geom_histogram()
```

```{r}
#One-Factor ANOVA
d3 <- d3 |> mutate(logRange.Size = log(Range.Size))
lm4 <- lm(logRange.Size ~ Migration, data = d3)
(summary(lm4))
```

Based on the global model, range size is significantly associated with Migration category. Looking at the R-squared value, 8.7% of the total variance in log(Range.Size) can be explained by Migration in this model. 

The **reference level** is **Migration category 1**. **Migration category 2** and **Migration category 3** both differ significantly from Migration category 1.

```{r}
#Releveling and assessing differences among the remaining pair of Migration categories.

d5 <- d |>
  mutate(Migration = relevel(Migration, ref = "3"))

lm5 <- lm(log(Mass) ~ Migration, data = d5)
(summary(lm5))
```

After releveling, **Migration category 3** is the **reference level**. Migration category 2 also differs significantly from Migration category 3.

```{r}
#ANOVA
maov2 <- aov(log(Range.Size) ~ Migration, data = d3)

#post-hoc Tukey Honest Significant Differences test
(posthoc <- TukeyHSD(maov2, which = "Migration", conf.level = 0.95))
```

Looking at the adjusted p-values, all three Migration categories significantly differ from each other. 

### Step 4

#### Run separate one-factor ANOVA analyses to look at the association between [1] relative beak length and Primary.Lifestyle and between [2] relative beak length and Trophic.Level. In doing so, [1] make boxplots of response variable by each predictor and by the combination of predictors, and [2] run linear models for each predictor separately and interpret the model output.

```{r}
#Winnowing data to just consider birds from the Order Passeriformes
d_pass <- d |> filter(Order1 == "Passeriformes")

#One-Factor ANOVA looking at association between relative beak length and Primary.Lifestyle
maov3 <- aov(relBeak.Length ~ Primary.Lifestyle, data = d_pass)
(summary(maov3))
```

```{r}
#Linear model looking at association between relative beak length and Primary.Lifestyle
lm6 <- lm(relBeak.Length ~ Primary.Lifestyle, data = d_pass)
(summary(lm6))
```

Based on the global model, relative beak length is significantly associated with Primary.Lifestyle. 5.6% of the variance in relative beak length can be explained by Primary.Lifesyle.

```{r}
#Boxplot looking at association between relative beak length and Primary.Lifestyle

p5 <- ggplot(d_pass |> filter(!is.na(Primary.Lifestyle)), aes(x = Primary.Lifestyle, y = relBeak.Length)) +
  geom_boxplot(na.rm = TRUE) +
  geom_jitter(alpha = 0.05) +
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("Relative Beak Length") + 
  xlab("Primary Lifestyle")
(p5)
```


```{r}
#One-Factor ANOVA looking at association between relative beak length and Trophic.Level
maov4 <- aov(relBeak.Length ~ Trophic.Level, data = d_pass)
(summary(maov4))
```

```{r}
#Linear model looking at association between relative beak length and Trophic.Level
lm7 <- lm(relBeak.Length ~ Trophic.Level, data = d_pass)
(summary(lm7))
```

Based on the global model, relative beak length is significantly associated with Trophic.Level. 5.0% of the variance in relative beak length can be explained by Trophic.Level. 

```{r}
#Boxplot looking at association between relative beak length and Trophic.Level

p6 <- ggplot(d_pass |> filter(!is.na(Trophic.Level)), aes(x = Trophic.Level, y = relBeak.Length)) +
  geom_boxplot(na.rm = TRUE) +
  geom_jitter(alpha = 0.05) +
  theme(axis.text.x = element_text(angle = 90)) + #Specifying axis labels
  ylab("Relative Beak Length") + 
  xlab("Trophic Level")
(p6)
```


### Step 5

#### Run a two-factor model to look at the association between relative beak length and both Primary.Lifestyle and Trophic.Level among the passeriforms

```{r}
#Two-Factor ANOVA without interaction term

maov5 <- aov(relBeak.Length ~ Trophic.Level + Primary.Lifestyle, data = d_pass)
(summary(maov5))
```

Holding Primary.Lifestyle constant, Trophic.Level has an impact on relative beak length. Holding Trophic.Level constant, Primary.Lifestyle has an impact on relative beak length.

### Step 6

#### Run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term.

```{r}
#Two-Factor ANOVA with interaction term
maov6 <- aov(relBeak.Length ~ Trophic.Level + Primary.Lifestyle + Trophic.Level:Primary.Lifestyle, data = d_pass)
(summary(maov6))
```

Based on the model output, I would conclude that the interaction between Trophic.Level and Primary.Lifestyle also is a significant predictor in variation in relative beak length.

### Step 7

#### Use the interaction.plot() function to visualize the interaction between Primary.Lifestyle and Trophic.Level

```{r}
p8 <- interaction.plot(x.factor = d_pass$Trophic.Level,
                 xlab = "Trophic Level",
                 trace.factor = d_pass$Primary.Lifestyle,
                 trace.label = "Primary Lifestyle",
                 response = d_pass$relBeak.Length, 
                 ylab = "Relative Beak Length")
(p8)
```

```{r}
#Flipping my axes
p9 <- interaction.plot(x.factor = d_pass$Primary.Lifestyle,
                 xlab = "Primary Lifestyle",
                 trace.factor = d_pass$Trophic.Level,
                 trace.label = "Trophic Level",
                 response = d_pass$relBeak.Length, 
                 ylab = "Relative Beak Length")
(p9)
```

